Simulating High-Dimensional Multivariate Data

### 2021-01-11 010-background.Rmd
2nd draft. Begin integrating new RNA-seq data

```{r ch010-myProb, include=FALSE, cache = TRUE}
myProb <- 0.95
```

```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
brcaRDS <- paste0( "data/brca", round(myProb*100), ".rds" )
if ( !file.exists( brcaRDS ) | !file.exists( './data/examplePatients.txt' ))  {
    ## full workflow simulating RNA-seq data
    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
    lastClinical <- which( names(allDat) == 'tumorsize' )
    brca <- allDat[ allDat$disease == "BRCA", ]
    ## Patient ID
    writeLines( as.character( brca$patient ), './data/examplePatients.txt' ) 
    brca <- brca[ , (lastClinical+1):ncol(allDat) ]
    dim(brca) ## num of genes 20501
    ## compute the median expression! avoid the 0s
    brcaMedian <- apply(brca, 2, median)
    ## retain top (1-myProb)*100% highest expressing genes for illustration
    cutPoint <- quantile( x = brcaMedian, probs = myProb )
    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    brca <- brca[ , genesToKeep ]
    ## ncol(brca) / 20501
    ## print(d <- length(genesToKeep))
    ## save data as rds
    saveRDS(brca, brcaRDS)
}
brca <- readRDS(file = brcaRDS)
d <- ncol(brca)
```

We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For simplicity, we only consider high expressing genes.
In turn, we begin by filtering to retain the top `r (1-myProb)*100`% of highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c RNA-seq by Expectation Maximization (RSEM) values to counts.
Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
if ( !file.exists( './data/exampleGenes.rds' ) )  {
    set.seed(10192020)
    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
    saveRDS(exampleGenes, file = './data/exampleGenes.rds')
}
exampleGenes <- readRDS('./data/exampleGenes.rds')
## Add row names per WWP comment
small <-  brca %>% select( exampleGenes ) %>% as.data.frame
numPatients <- 5
smallForTable <- head( small, n = numPatients )
smallForTable$PatientID <- head( readLines('./data/examplePatients.txt'), n = numPatients )
smallForTable <- smallForTable[ , c(4, 1:3 ) ]
smallForTable[ , 2:4 ] <- round( smallForTable[ , 2:4 ] )
knitr::kable(smallForTable ,
      booktabs = TRUE,
      caption = "mRNA expression for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
)
```

>>>>

```{r ch010-myD, include=FALSE, cache = TRUE, eval=T}
d <- 1000
```

```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
brcaRDS <- paste0( "data/brca", d, ".rds" )
if ( !file.exists( brcaRDS ) | !file.exists( './data/examplePatients.txt' ) )  {
    brca0 <- TCGA2STAT::getTCGA(disease = "BRCA",
                            data.type = "RNASeq",
                            p = 6L)
    brca1 <- as.data.frame(t(brca0$dat))
    gene_median <- apply(brca1, 2, median)
    gene_median_order <- order(gene_median, decreasing = TRUE)
    brca  <- brca1[,head(gene_median_order, d)]
    writeLines( substring( row.names( brca ), first=1, last= 12 ), './data/examplePatients.txt' ) 
    saveRDS(brca, brcaRDS)
}
brca <- readRDS(file = brcaRDS)
```

We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For simplicity, we only consider high expressing genes.
In turn, we begin by filtering to retain the top `r d`% of highest-expressing genes (in terms of median expression) of the 20,502 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples.
This gives a great number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
if ( !file.exists( './data/exampleGenes.rds' ) )  {
    set.seed(01112021)
    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
    saveRDS(exampleGenes, file = './data/exampleGenes.rds')
}
exampleGenes <- readRDS('./data/exampleGenes.rds')
## Add row names per WWP comment
small <-  brca %>% select( exampleGenes ) %>% as.data.frame
numPatients <- 5
smallForTable <- head( small, n = numPatients )
smallForTable$PatientID <- head( readLines('./data/examplePatients.txt'), n = numPatients )
smallForTable <- smallForTable[ , c(4, 1:3 ) ]
smallForTable[ , 2:4 ] <- round( smallForTable[ , 2:4 ] )
knitr::kable(smallForTable ,
      booktabs = TRUE,
      caption = "mRNA expression for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
)
```

### 2021-01-11 index.Rmd
2nd draft. Begin integrating new RNA-seq data

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
site: bookdown::bookdown_site
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and evaluating quantitative methods. But measurements are often correlated and high dimensional in this era of big data, however, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and a broad class of admissible Spearman correlation and Kendall $\tau$ matrices. We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d=1026$.
---

>>>>

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
site: bookdown::bookdown_site
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and evaluating quantitative methods. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and a broad class of admissible Spearman correlation and Kendall $\tau$ matrices. We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
---
